### 2022-12-24 agda/LRKits.lagda.tex
[ edit ] another draft of chapters 5 and 6

  open import Algebra.Relational renaming (_↘_↙_ to _↘-_-↙_)
  open import Data.LTree
  open import Data.LTree.Vector
  open import Data.Product
  open import Data.Wrap
  open import Function using (id; _∘_; Equivalence)
  open import Relation.Binary.PropositionalEquality as ≡ using (_≡_)
  open import Relation.Nary
  open import Relation.Unary.Bunched

@@

  open import Generic.Linear.Operations rawPoSemiring
  open import Generic.Linear.Algebra poSemiring
  open import Generic.Linear.Syntax Ty Ann
  open import Generic.Linear.Syntax.Interpretation Ty rawPoSemiring
  open import Generic.Linear.Variable Ty rawPoSemiring hiding (_↙ᵛ_; _↘ᵛ_)
  open import Generic.Linear.Environment Ty poSemiring
  open import Generic.Linear.Environment.Properties Ty poSemiring
  open import Generic.Linear.Environment.Categorical Ty poSemiring
\end{code}

@@

  data _⊢_ : Ctx → Ty → Set where
    var : ∀[ _∋_ ⇒ _⊢_ ]
    !-I : ∀ {r A} → ∀[ r ·ᶜ (_⊢ A) ⇒ _⊢ `! r A ]
    !-E : ∀ {r A C} → ∀[ (_⊢ `! r A) ✴ᶜ Bind [ r · A ]ᶜ (_⊢ C) ⇒ _⊢ C ]
\end{code}
%</Tm>

@@

    env-+ : ∀ {s t γ δ P Qₗ Qᵣ} →
      let Γ = ctx {s} P γ; Δₗ = ctx {t} Qₗ δ; Δᵣ = ctx {t} Qᵣ δ in
      (λ Q → [ 𝓥 ] Γ ⇒ᵉ ctx Q δ) ◇ _≤[ Qₗ +* Qᵣ ] →
      (λ Pₗ → [ 𝓥 ] ctx Pₗ γ ⇒ᵉ Δₗ)
        ↘- P ≤[_+*_] -↙
      (λ Pᵣ → [ 𝓥 ] ctx Pᵣ γ ⇒ᵉ Δᵣ)
    env-+ (ρ , sp) =
      let fit-hereₗ ↘, sp′ ,↙ fit-hereᵣ = ρ .Ψ .rel-+ₘ (sp , ρ .fit-here) in
      record { [_]_⇒ᵉ_ ρ; fit-here = fit-hereₗ }
        ↘, sp′ ,↙
      record { [_]_⇒ᵉ_ ρ; fit-here = fit-hereᵣ }

@@

%<*Kit>
\begin{code}
  Weakening : ∀ {v} {A : Set} → (Ctx → A → Set v) → Set v
  Weakening 𝓥 =
    ∀ {s t P} {γ : Vector Ty (s <+> t)} → P ∘ ↘ ≤* 0* →
    ∀[ 𝓥 (ctx (P ∘ ↙) (γ ∘ ↙)) ⇒ 𝓥 (ctx P γ) ]

  record Kit (𝓥 : Ctx → Ty → Set) : Set where
    constructor kit
    field
      ↙ᵏ : Weakening 𝓥
      vr : ∀[ _∋_ ⇒ 𝓥 ]
      tm : ∀[ 𝓥 ⇒ _⊢_ ]
\end{code}
%</Kit>

@@

%<*bindEnv>
\begin{code}
    bindEnv : ∀ {Γ Δ Θ} → [ 𝓥 ] Γ ⇒ᵉ Δ → [ 𝓥 ] Γ ++ᶜ Θ ⇒ᵉ Δ ++ᶜ Θ
    bindEnv ρ .Ψ = [ [ ρ .Ψ │ 0ᴿ ]ᴿ ─ [ 0ᴿ │ 1ᴿ ]ᴿ ]ᴿ
    bindEnv ρ .fit-here = _↘,_,↙_
      {left = _ ++ _} {right = _ ++ _}
      (ρ .fit-here , 0*-triv)
      (+*-identity↘ _ ++ₙ +*-identity↙ _)
      (0*-triv , ≤*-refl)
    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar i q b)
      with un++ₙ sp+ | un++ₙ b
    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar (↙ i) q b)
      | sp+l , sp+r | bl , br = ↙ᵏ
      (≤*-trans (+ₘ-identityˡ→ (ur , sp+r)) (≤*-trans dr br))
      (ρ .lookup (ρ .Ψ .rel-≤ₘ ≤*-refl (+ₘ-identityʳ→ (sp+l , dl)) ul)
        (lvar i q bl))
    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar (↘ i) q b)
      | sp+l , sp+r | bl , br = vr (lvar (↘ i) q
      (≤*-trans (+ₘ-identityʳ→ (sp+l , dl))
        (0*→≤* (ρ .Ψ .rel-0ₘ (≤*→0* bl , ul)))
      ++ₙ
      ≤*-trans (+ₘ-identityˡ→ (ur , sp+r)) (≤*-trans dr br)))
\end{code}
%</bindEnv>

%<*trav>
\begin{code}
    trav : ∀ {Γ Δ A} → [ 𝓥 ] Γ ⇒ᵉ Δ → Δ ⊢ A → Γ ⊢ A
    trav ρ (var x) = tm (ρ .lookup (ρ .fit-here) x)
    trav ρ (!-I (⟨ sp* ⟩· M)) =
      let sp*′ , ρ′ = env-* (ρ , sp*) in
      !-I (⟨ sp*′ ⟩· (trav ρ′ M))
    trav ρ (!-E (M ✴⟨ sp+ ⟩ N)) =
      let ρₗ ↘, sp+′ ,↙ ρᵣ = env-+ (ρ , sp+) in
      !-E (trav ρₗ M ✴⟨ sp+′ ⟩ trav (bindEnv ρᵣ) N)
\end{code}
%</trav>

@@

  ren = trav ∋-kit
\end{code}
%</var-kit>

@@

  sub = trav ⊢-kit
\end{code}
%</sub>

@@

%<*subSingle>
\begin{code}
    sub[-] : ∀ {s P Q R γ r A B} → R ≤[ Q +*_] ◇ _≤[ r *ₗ P ] →
      ctx {s} P γ ⊢ A → ctx Q γ ++ᶜ [ r · A ]ᶜ ⊢ B → ctx R γ ⊢ B
    sub[-] {Q = Q} {R} {γ} {r} {A} (sp+ , sp*) t = sub σ
      where
      σ : [ _⊢_ ] ctx R γ ⇒ᵉ ctx Q γ ++ᶜ [ r · A ]ᶜ
      σ = ++ᵉ (id^Env ✴⟨ sp+ ⟩ [-]ᵉ (⟨ sp* ⟩· t))
\end{code}
%</subSingle>

>>>>

  open import Algebra.Relational renaming (_↘_↙_ to _↘-_-↙_)
  open import Data.LTree
  open import Data.LTree.Vector
  open import Data.Product
  open import Data.Wrap
  open import Function using (id; _∘_; Equivalence)
  open import Relation.Binary.PropositionalEquality as ≡ using (_≡_)
  open import Relation.Nary.Extra using (_⊆_)
  open import Relation.Unary.Bunched

@@

  open import Generic.Linear.Operations rawPoSemiring
  open import Generic.Linear.Algebra poSemiring
  open import Generic.Linear.Syntax Ty Ann
  open import Generic.Linear.Syntax.Interpretation Ty rawPoSemiring hiding (◇)
  open import Generic.Linear.Variable Ty rawPoSemiring hiding (_↙ᵛ_; _↘ᵛ_)
  open import Generic.Linear.Environment Ty poSemiring
  open import Generic.Linear.Environment.Properties Ty poSemiring
  open import Generic.Linear.Environment.Categorical Ty poSemiring
\end{code}

@@

  data _⊢_ : Ctx → Ty → Set where
    var : _∋_ ⊆ _⊢_
    !-I : ∀ {r A} → r · (_⊢ A) ⊆ _⊢ `! r A
    !-E : ∀ {r A C} → (_⊢ `! r A) ✴ Bind [ r ∙ A ]ᶜ (_⊢ C) ⊆ _⊢ C
\end{code}
%</Tm>

@@

    env-+ : ∀ {s t γ δ P Qₗ Qᵣ} →
      let Γ = ctx {s} P γ; Δₗ = ctx {t} Qₗ δ; Δᵣ = ctx {t} Qᵣ δ in
      (λ Q → [ 𝓥 ] Γ ⇒ᵉ ctx Q δ) ◇ (_≤[ Qₗ +* Qᵣ ]) →
      (λ Pₗ → [ 𝓥 ] ctx Pₗ γ ⇒ᵉ Δₗ)
        ↘- P ≤[_+*_] -↙
      (λ Pᵣ → [ 𝓥 ] ctx Pᵣ γ ⇒ᵉ Δᵣ)
    env-+ (ρ , sp) =
      let fit-hereₗ ↘, sp′ ,↙ fit-hereᵣ = ρ .Ψ .rel-+ₘ (sp , ρ .fit-here) in
      record { [_]_⇒ᵉ_ ρ; fit-here = fit-hereₗ }
        ↘, sp′ ,↙
      record { [_]_⇒ᵉ_ ρ; fit-here = fit-hereᵣ }

@@

%<*Kit>
\begin{code}
  Weakening : ∀ {v} {A : Set} → (Ctx → A → Set v) → Set v
  Weakening 𝓥 =
    ∀ {s t P} {γ : Vector Ty (s <+> t)} → P ∘ ↘ ≤* 0* →
    𝓥 (ctx (P ∘ ↙) (γ ∘ ↙)) ⊆ 𝓥 (ctx P γ)

  record Kit (𝓥 : Ctx → Ty → Set) : Set where
    constructor kit
    field
      ↙ᵏ : Weakening 𝓥
      vr : _∋_ ⊆ 𝓥
      tm : 𝓥 ⊆ _⊢_
\end{code}
%</Kit>

@@

%<*bindEnv>
\begin{code}
    bindEnv : ∀ {Γ Δ Θ} → [ 𝓥 ] Γ ⇒ᵉ Δ → [ 𝓥 ] Γ ++ᶜ Θ ⇒ᵉ Δ ++ᶜ Θ
\end{code}
%</bindEnv>
\begin{code}
    bindEnv ρ .Ψ = [  [ ρ .Ψ  │ 0ᴿ ]ᴿ
                              ─
                      [ 0ᴿ    │ 1ᴿ ]ᴿ ]ᴿ
    bindEnv ρ .fit-here = _↘,_,↙_
      {left = _ ++ _} {right = _ ++ _}
      (ρ .fit-here , 0*-triv)
      (+*-identity↘ _ ++ₙ +*-identity↙ _)
      (0*-triv , ≤*-refl)
    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar i q b)
      with un++ₙ sp+ | un++ₙ b
    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar (↙ i) q b)
      | sp+l , sp+r | bl , br = ↙ᵏ
      (≤*-trans (+ₘ-identityˡ→ (ur , sp+r)) (≤*-trans dr br))
      (ρ .lookup (ρ .Ψ .rel-≤ₘ ≤*-refl (+ₘ-identityʳ→ (sp+l , dl)) ul)
        (lvar i q bl))
    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar (↘ i) q b)
      | sp+l , sp+r | bl , br = vr (lvar (↘ i) q
      (≤*-trans (+ₘ-identityʳ→ (sp+l , dl))
        (0*→≤* (ρ .Ψ .rel-0ₘ (≤*→0* bl , ul)))
      ++ₙ
      ≤*-trans (+ₘ-identityˡ→ (ur , sp+r)) (≤*-trans dr br)))
\end{code}

%<*trav>
\begin{code}
    trav : ∀ {Γ Δ A} → [ 𝓥 ] Γ ⇒ᵉ Δ → Δ ⊢ A → Γ ⊢ A
    trav ρ (var x) = tm (ρ .lookup (ρ .fit-here) x)
    trav ρ (!-I (⟨ sp* ⟩·ᶜ M)) =
      let sp*′ , ρ′ = env-* (ρ , sp*) in
      !-I (⟨ sp*′ ⟩·ᶜ (trav ρ′ M))
    trav ρ (!-E (M ✴ᶜ⟨ sp+ ⟩ N)) =
      let ρₗ ↘, sp+′ ,↙ ρᵣ = env-+ (ρ , sp+) in
      !-E (trav ρₗ M ✴ᶜ⟨ sp+′ ⟩ trav (bindEnv ρᵣ) N)
\end{code}
%</trav>

@@

  ren : ∀ {Γ Δ A} → [ _∋_ ] Γ ⇒ᵉ Δ → Δ ⊢ A → Γ ⊢ A
  ren = trav ∋-kit
\end{code}
%</var-kit>

@@

  sub : ∀ {Γ Δ A} → [ _⊢_ ] Γ ⇒ᵉ Δ → Δ ⊢ A → Γ ⊢ A
  sub = trav ⊢-kit
\end{code}
%</sub>

@@

%<*subSingle>
\begin{code}
    sub[-] : ∀ {s P Q R γ r A B} → R ≤[ Q +*_] ◇ _≤[ r *ₗ P ] →
      ctx {s} P γ ⊢ A → ctx Q γ ++ᶜ [ r ∙ A ]ᶜ ⊢ B → ctx R γ ⊢ B
    sub[-] {Q = Q} {R} {γ} {r} {A} (sp+ , sp*) t = sub σ
      where
      σ : [ _⊢_ ] ctx R γ ⇒ᵉ ctx Q γ ++ᶜ [ r ∙ A ]ᶜ
      σ = ++ᵉ (id^Env ✴ᶜ⟨ sp+ ⟩ [-]ᵉ (⟨ sp* ⟩·ᶜ t))
\end{code}
%</subSingle>

### 2022-12-24 tex/examples.tex
[ writing ] content sections to full draft
[ edit ] another draft of chapters 5 and 6

\subsection{Renaming and substitution}\label{sec:kits}

@@

@@

\todo{Define big intersection.}
When the poset of worlds forms a (relational) commutative monoid, such
world-indexed relations support a symmetric monoidal closed structure, with
objects denoted \AgdaFunction{I$^R$},
\AgdaFunction{\AgdaUnderscore{}$\otimes^R$\AgdaUnderscore{}}, and
\AgdaFunction{\AgdaUnderscore{}$\multimap^R$\AgdaUnderscore{}},.
These reuse the bunched connectives \AgdaRecord{$I^*$}, \AgdaRecord{$\sep$}, and
\AgdaRecord{$\wand$}, now over worlds rather than contexts.

>>>>

\subsection{Renaming and substitution}\label{sec:kit-to-sem}

@@

Standard denotational semantics falls out as a somewhat trivial case of
\AgdaRecord{Semantics}.
A lot of work is done in the generic traversal \AgdaFunction{semantics} to
maintain a $\V$-environment, where $\V$ is a reasonably variable-like semantic
family.

{\color{red}Old stuff below\ldots}

@@

When the poset of worlds forms a (relational) commutative monoid, such
world-indexed relations support a symmetric monoidal closed structure, with
objects denoted \AgdaFunction{I$^R$},
\AgdaFunction{\AgdaUnderscore{}$\otimes^R$\AgdaUnderscore{}}, and
\AgdaFunction{\AgdaUnderscore{}$\multimap^R$\AgdaUnderscore{}},.
These reuse the bunched connectives \AgdaRecord{$I^*$}, \AgdaRecord{$\sep$}, and
\AgdaRecord{$\wand$}, now over worlds rather than contexts.

### 2022-12-24 tex/lenv.tex
[ edit ] another draft of chapters 5 and 6

  \forallb{\V\,\grP\gamma \dotto \V\,\grPprime\gamma}$).

@@

    \item $\forallb{I \dotlr \plr{{-} \env\V {\cdot}}}$
    \item $\forallb{\plr{{-} \env\V \Delta_l} \sep \plr{{-} \env\V \Delta_r}
      \dotlr \plr{{-} \env\V \Delta_l, \Delta_r}}$
    \item
      $\forallb{\gr r \cdot \plr{\V\,(-)\,A} \dotlr \plr{{-} \env\V \gr rA}}$

@@

    \item[$I(\to)$]

@@

    \item[$I(\gets)$]

@@

    \item[$\sep(\to)$]

@@

    \item[$\sep(\gets)$]

@@

    \item[$\cdot(\to)$]

@@

    \item[$\cdot(\gets)$]

@@

\begin{lemma}[\AgdaFunction{bindEnv}]\label{thm:lr-bind}
  Given functions
  ${\swarrow^k} : \forall \Gamma, \grR, \theta.~\grR \leq \gr0 \to
  \forallb{\V\,\Gamma \dotto \V\,\plr{\Gamma, \grR\theta}}$ and
  $\mathrm{vr} : \forallb{{\sqni} \dotto \V}$, we can turn an environment of
  type $\Gamma \env\V \Delta$ into an environment of type
  $\Gamma, \Theta \env\V \Delta, \Theta$ for any context $\Theta$.
\end{lemma}
\begin{proof}
  Let $\grP\gamma \coloneqq \Gamma$, $\grQ\delta \coloneqq \Delta$, and
  $\grR\theta \coloneqq \Theta$.
  Let the new linear map $\gr\Psi\gr' : \Ann^{\size\Delta + \size\Theta} \to
  \Ann^{\size\Gamma + \size\Theta}$ be $\gr\Psi \oplus \gr I$.
  That is, in block matrix notation,
  $\begin{pmatrix} \gr\Psi & \gr0 \\ \gr0 & \gr I \end{pmatrix}$.
  Checking that this linear map fits, we have
  $\begin{pmatrix}\grP & \grR\end{pmatrix}
  \leq \begin{pmatrix}\grQ\gr\Psi & \grR\gr I\end{pmatrix}
  = \begin{pmatrix}\grQ & \grR\end{pmatrix}\plr{\gr\Psi \oplus \gr I}$.
  For the action on variables, we are given vectors $\grPprime$,
  $\grR\gr'_\grP$, $\grQprime$, and $\grR\gr'_\grQ$ such that
  $\begin{pmatrix} \grPprime & \grR\gr'_\grP \end{pmatrix} \leq
  \begin{pmatrix} \grQprime & \grR\gr'_\grQ \end{pmatrix}
  \plr{\gr\Psi \oplus \gr I}$ and we have a variable of type
  $\grQprime\delta, \grR\gr'_\grQ\theta \sqni A$ for some type $A$.
  The constraint on the new vectors reduces to $\grPprime \leq \grQprime\gr\Psi$
  and $\grR\gr'_\grP \leq \grR\gr'_\grQ$.
  From the variable we either have a variable $x$ in $\delta$ with
  $\grQprime \leq \langle x \rvert$ and $\grR\gr'_\grQ \leq \gr0$, or a
  variable $y$ in $\theta$ with $\grQprime \leq \gr0$ and
  $\grR\gr'_\grQ \leq \langle y \rvert$.
  In the former case, the action of the original environment on $x$ gives us a
  $\V$-value in $\grPprime\gamma$, and the $\gr0$-weakening principle
  $\swarrow^k$, noting that $\grR\gr'_\grP \leq \grR\gr'_\grQ \leq \gr0$, gives
  us a $\V$-value in $\grPprime\gamma, \grR\gr'_\grP\theta$.
  In the latter case, we have that
  $\begin{pmatrix} \grPprime & \grR\gr'_\grP \end{pmatrix}
  \leq \begin{pmatrix} \grQprime\gr\Psi & \grR\gr'_\grQ \end{pmatrix}
  \leq \begin{pmatrix} \gr0\gr\Psi & \langle y \rvert \end{pmatrix}
  = \begin{pmatrix} \gr0 & \langle y \rvert \end{pmatrix}
  = \left\langle {\searrow}y \right\rvert$, so $y$ also serves as a
  usage-checked variable in $\grPprime\gamma, \grR\gr'_\grP\theta$.
  From this usage-checked variable, we get a $\V$-value in the same context
  using $\mathrm{vr}$.
\end{proof}

@@

Following \citet{ACU15}, we expect (intuitionistic) ST$\lambda$C syntax to form
a relative monad over $\ni$ seen as a functor from the category of contexts
under renaming to the functor category $\blr{\mathrm{Ty}, \Set}$, where
$\mathrm{Ty}$ is the discrete category of ST$\lambda$C types.
Notice that, given $F,G : \blr{\mathrm{Ty}, \Set}$, a morphism from $F$ to $G$
is a function of type $\forallb{F \dotto G}$ (with naturality being trivial).
Therefore, we expect a relative monad, given as a Kleisli triple, to have a unit
$\eta_\Gamma : \forallb{\Gamma \ni \plr{-} \dotto \Gamma \vdash \plr{-}}$
given by the variable rule, and a Kleisli extension operator
${^*}_{\Gamma,\Delta} :
\forallb{\Gamma \ni \plr{-} \dotto \Delta \vdash \plr{-}} \to
\forallb{\Gamma \vdash \plr{-} \dotto \Delta \vdash \plr{-}}$
given by substitution.
Composition of substitutions falls out of this framework as Kleisli composition.
However, in the usage-aware case, substitution needs not just a mapping of
variables $f : \forallb{\Gamma \sqni \plr{-} \dotto \Delta \vdash \plr{-}}$,
but rather an environment $\rho : \Delta \env\vdash \Gamma$, as we have already
discussed.
It therefore makes sense for our replacement for the Kleisli extension operator
to similarly take an environment rather than a simple variable mapping.

\Cref{thm:env-comp-lemma} below amounts to deriving a modified notion of Kleisli
composition from a modified Kleisli extension.
Additionally, \cref{thm:env-id-lemma} is required to turn a monadic unit into an
identity environment.
Both lemmas are stated in terms of general $\U$/$\V$/$\W$-environments, with
some specific examples (e.g.\ for renaming and substitution) below them.

\begin{lemma}[Identity environment]\label{thm:env-id-lemma}
  Given a function
  \[
    \mathrm{vr}_{\Gamma'} : \forallb{\Gamma' \sqni \plr{-} \dotto \V\,\Gamma'}
  \]
  for any
  $\Gamma$ we have an environment $\mathrm{id} : \Gamma \env\V \Gamma$.
\end{lemma}
\begin{proof}
  Let $\Gamma = \grP\gamma$.
  Let $\gr\Psi$ be the identity map, which clearly satisfies
  $\grP \leq \grP\gr\Psi$.
  When acting on a variable, the inequality $\grPprime \leq \grQprime\gr\Psi$
  means that $\grPprime \leq \grQprime$.
  We are given a variable of type $\grQprime\gamma \sqni A$, which we can
  coerce to a variable of type $\grPprime\gamma \sqni A$, upon which we apply
  $\mathrm{vr}$ to get the required value of type $\V\,\grPprime\gamma\,A$.
\end{proof}

\begin{lemma}[Composition of environments]\label{thm:env-comp-lemma}
  Given a function
  \[
    \mathrm{lift}_{\Gamma', \Delta'} :
    \Gamma' \env\U \Delta' \to \forallb{\V\,\Delta' \dotto \W\,\Gamma'}
  \]
  we can compose environments $\rho : \Gamma \env\U \Delta$ and
  $\sigma : \Delta \env\V \Theta$ into an environment
  $\rho \gg \sigma : \Gamma \env\W \Theta$.
\end{lemma}
\begin{proof}
  Let $\Gamma = \grP\gamma$, $\Delta = \grQ\delta$, and $\Theta = \grR\theta$.
  Take $\gr\Psi$ to be the composition $\plr{\sigma.\gr\Psi}\plr{\rho.\gr\Psi}$,
  noting that
  $\grP \leq \grQ\plr{\rho.\gr\Psi} \leq
  (\grR\plr{\sigma.\gr\Psi})\plr{\rho.\gr\Psi} = \grR\gr\Psi$
  thanks to the inequalities yielded by $\sigma$ and $\rho$.
  When acting on a variable, we are given $\grPprime \leq \grRprime\gr\Psi$ and
  a variable $v : \grRprime\theta \sqni A$, and want a value of type
  $\W\,\grPprime\gamma\,A$.
  Let $\grQprime \coloneqq \grRprime\plr{\sigma.\gr\Psi}$, with inequality
  $\grQprime \leq \grRprime\plr{\sigma.\gr\Psi}$ giving us a value
  $\sigma(v) : \V\,\grQprime\delta\,A$.
  We wish to apply $\mathrm{lift}$ to $\sigma(v)$ with
  $\Gamma' \coloneqq \grPprime\gamma$ and $\Delta' \coloneqq \grQprime\delta$ to
  complete the construction of the $\W$-value.
  To do this, we need an environment of type
  $\grPprime\gamma \env\U \grQprime\delta$, which we can get from $\rho$ using
  \cref{thm:env-resize}, noting that
  $\grPprime \leq \grRprime\plr{\sigma.\gr\Psi}\plr{\rho.\gr\Psi} =
  \grQprime\plr{\rho.\gr\Psi}$.
\end{proof}

>>>>

  \V\,\grP\gamma \rightarrowtriangle \V\,\grPprime\gamma$).

@@

    \item $I^{\sep} \leftrightarrowtriangle \plr{{-} \env\V {\cdot}}$
    \item $\plr{{-} \env\V \Delta_l} \sep \plr{{-} \env\V \Delta_r}
      \leftrightarrowtriangle \plr{{-} \env\V \Delta_l, \Delta_r}$
    \item $\gr r \cdot \plr{\V\,(-)\,A}
      \leftrightarrowtriangle \plr{{-} \env\V \gr rA}$

@@

    \item[$I^{\sep}(\rightarrowtriangle)$]

@@

    \item[$I^{\sep}(\leftarrowtriangle)$]

@@

    \item[$\sep(\rightarrowtriangle)$]

@@

    \item[$\sep(\leftarrowtriangle)$]

@@

    \item[$\cdot(\rightarrowtriangle)$]

@@

    \item[$\cdot(\leftarrowtriangle)$]

@@

\begin{lemma}[\AgdaFunction{bindEnv}]\label{thm:lr-bind}
  Given functions
  ${\swarrow^k} : \forall \Gamma, \grR, \theta.~\grR \leq \gr0 \to
  \V\,\Gamma \rightarrowtriangle \V\,\plr{\Gamma, \grR\theta}$ and
  $\mathrm{vr} : {\sqni} \rightarrowtriangle \V$, we can turn an environment of
  type $\Gamma \env\V \Delta$ into an environment of type
  $\Gamma, \Theta \env\V \Delta, \Theta$ for any context $\Theta$.
\end{lemma}
\begin{proof}
  Let $\grP\gamma \coloneqq \Gamma$, $\grQ\delta \coloneqq \Delta$, and
  $\grR\theta \coloneqq \Theta$.
  Let the new linear map $\gr\Psi\gr' : \Ann^{\size\Delta + \size\Theta} \to
  \Ann^{\size\Gamma + \size\Theta}$ be $\gr\Psi \oplus \gr I$.
  That is, in block matrix notation,
  $\begin{pmatrix} \gr\Psi & \gr0 \\ \gr0 & \gr I \end{pmatrix}$.
  Checking that this linear map fits, we have
  $\begin{pmatrix}\grP & \grR\end{pmatrix}
  \leq \begin{pmatrix}\grQ\gr\Psi & \grR\gr I\end{pmatrix}
  = \begin{pmatrix}\grQ & \grR\end{pmatrix}\plr{\gr\Psi \oplus \gr I}$.
  For the action on variables, we are given vectors $\grPprime$,
  $\grR\gr'_\grP$, $\grQprime$, and $\grR\gr'_\grQ$ such that
  $\begin{pmatrix} \grPprime & \grR\gr'_\grP \end{pmatrix} \leq
  \begin{pmatrix} \grQprime & \grR\gr'_\grQ \end{pmatrix}
  \plr{\gr\Psi \oplus \gr I}$ and we have a variable of type
  $\grQprime\delta, \grR\gr'_\grQ\theta \sqni A$ for some type $A$.
  The constraint on the new vectors reduces to $\grPprime \leq \grQprime\gr\Psi$
  and $\grR\gr'_\grP \leq \grR\gr'_\grQ$.
  From the variable we either have a variable $x$ in $\delta$ with
  $\grQprime \leq \langle x \rvert$ and $\grR\gr'_\grQ \leq \gr0$, or a
  variable $y$ in $\theta$ with $\grQprime \leq \gr0$ and
  $\grR\gr'_\grQ \leq \langle y \rvert$.
  In the former case, the action of the original environment on $x$ gives us a
  $\V$-value in $\grPprime\gamma$, and the $\gr0$-weakening principle
  $\swarrow^k$, noting that $\grR\gr'_\grP \leq \grR\gr'_\grQ \leq \gr0$, gives
  us a $\V$-value in $\grPprime\gamma, \grR\gr'_\grP\theta$.
  In the latter case, we have that
  $\begin{pmatrix} \grPprime & \grR\gr'_\grP \end{pmatrix}
  \leq \begin{pmatrix} \grQprime\gr\Psi & \grR\gr'_\grQ \end{pmatrix}
  \leq \begin{pmatrix} \gr0\gr\Psi & \langle y \rvert \end{pmatrix}
  = \begin{pmatrix} \gr0 & \langle y \rvert \end{pmatrix}
  = \left\langle {\searrow}y \right\rvert$, so $y$ also serves as a
  usage-checked variable in $\grPprime\gamma, \grR\gr'_\grP\theta$.
  From this usage-checked variable, we get a $\V$-value in the same context
  using $\mathrm{vr}$.
\end{proof}

@@

Following \citet{ACU15}, we expect (intuitionistic) ST$\lambda$C syntax to form
a relative monad over $\ni$ seen as a functor from the category of contexts
under renaming to the functor category $\blr{\mathrm{Ty}, \Set}$, where
$\mathrm{Ty}$ is the discrete category of ST$\lambda$C types.
Notice that, given $F,G : \blr{\mathrm{Ty}, \Set}$, a morphism from $F$ to $G$
is a function of type $F \rightarrowtriangle G$ (with naturality being trivial).
Therefore, we expect a relative monad, given as a Kleisli triple, to have a unit
$\eta_\Gamma : \Gamma \ni \plr{-} \rightarrowtriangle \Gamma \vdash \plr{-}$
given by the variable rule, and a Kleisli extension operator
${^*}_{\Gamma,\Delta} :
\plr{\Gamma \ni \plr{-} \rightarrowtriangle \Delta \vdash \plr{-}} \to
\plr{\Gamma \vdash \plr{-} \rightarrowtriangle \Delta \vdash \plr{-}}$
given by substitution.
Composition of substitutions falls out of this framework as Kleisli composition.
However, in the usage-aware case, substitution needs not just a mapping of
variables $f : \Gamma \sqni \plr{-} \rightarrowtriangle \Delta \vdash \plr{-}$,
but rather an environment $\rho : \Delta \env\vdash \Gamma$, as we have already
discussed.
It therefore makes sense for our replacement for the Kleisli extension operator
to similarly take an environment rather than a simple variable mapping.

\Cref{thm:env-comp} below amounts to deriving a modified notion of Kleisli
composition from a modified Kleisli extension.
Additionally, \cref{thm:env-id} is required to turn a monadic unit into an
identity environment.
Both lemmas are stated in terms of general $\U$/$\V$/$\W$-environments, with
some specific examples (e.g.\ for renaming and substitution) below them.

\begin{lemma}[Identity environment]\label{thm:env-id}
  Given a function
  \[
    \mathrm{vr}_{\Gamma'} :
    \Gamma' \sqni \plr{-} \rightarrowtriangle \V\,\Gamma'
  \]
  for any
  $\Gamma$ we have an environment $\mathrm{id} : \Gamma \env\V \Gamma$.
\end{lemma}
\begin{proof}
  Let $\Gamma = \grP\gamma$.
  Let $\gr\Psi$ be the identity map, which clearly satisfies
  $\grP \leq \grP\gr\Psi$.
  When acting on a variable, the inequality $\grPprime \leq \grQprime\gr\Psi$
  means that $\grPprime \leq \grQprime$.
  We are given a variable of type $\grQprime\gamma \sqni A$, which we can
  coerce to a variable of type $\grPprime\gamma \sqni A$, upon which we apply
  $\mathrm{vr}$ to get the required value of type $\V\,\grPprime\gamma\,A$.
\end{proof}

\begin{lemma}[Composition of environments]\label{thm:env-comp}
  Given a function
  \[
    \mathrm{lift}_{\Gamma', \Delta'} :
    \Gamma' \env\U \Delta' \to \V\,\Delta' \rightarrowtriangle \W\,\Gamma'
  \]
  we can compose environments $\rho : \Gamma \env\U \Delta$ and
  $\sigma : \Delta \env\V \Theta$ into an environment
  $\rho \gg \sigma : \Gamma \env\W \Theta$.
\end{lemma}
\begin{proof}
  Let $\Gamma = \grP\gamma$, $\Delta = \grQ\delta$, and $\Theta = \grR\theta$.
  Take $\gr\Psi$ to be the composition $\plr{\sigma.\gr\Psi}\plr{\rho.\gr\Psi}$,
  noting that
  $\grP \leq \grQ\plr{\rho.\gr\Psi} \leq
  (\grR\plr{\sigma.\gr\Psi})\plr{\rho.\gr\Psi} = \grR\gr\Psi$
  thanks to the inequalities yielded by $\sigma$ and $\rho$.
  When acting on a variable, we are given $\grPprime \leq \grRprime\gr\Psi$ and
  a variable $v : \grRprime\theta \sqni A$, and want a value of type
  $\W\,\grPprime\gamma\,A$.
  Let $\grQprime \coloneqq \grRprime\plr{\sigma.\gr\Psi}$, with inequality
  $\grQprime \leq \grRprime\plr{\sigma.\gr\Psi}$ giving us a value
  $\sigma(v) : \V\,\grQprime\delta\,A$.
  We wish to apply $\mathrm{lift}$ to $\sigma(v)$ with
  $\Gamma' \coloneqq \grPprime\gamma$ and $\Delta' \coloneqq \grQprime\delta$ to
  complete the construction of the $\W$-value.
  To do this, we need an environment of type
  $\grPprime\gamma \env\U \grQprime\delta$, which we can get from $\rho$ using
  \cref{thm:env-resize}, noting that
  $\grPprime \leq \grRprime\plr{\sigma.\gr\Psi}\plr{\rho.\gr\Psi} =
  \grQprime\plr{\rho.\gr\Psi}$.
\end{proof}
